Suitability prevalence area index in late quaternary explains genetic diversity

### 2023-07-11 manuscript_english.Rmd
change abstract

## Abstract
As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].

Our results reveal that fitness prevalence corresponds to the fixation index of *S. aberti* populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value. Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.

@@

@@

For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$. We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$. Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions. If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$. i. e. the set of environmental conditions for which values of fitness are capable of sustaining populations. We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$. This is a way of expressing the fundamental niche as a function of environmental conditions. So we can map each point of geography to ambient space. and, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$. Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.

@@

The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

@@

![Hypothetical niche in environmental space. An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}

@@

Moreover, the effective population can be estimated as the harmonic mean of the population in a demographic time series [@karlin1968rates] for $\tau$ generations $\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)$. We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close.

$$
F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_j} 
$$

@@

We call this harmonic mean of environmental suitability *Suitability prevalence index *($S$) given between $0$ and $1$, since it reflects the rate of population growth as a function of environment over a period of time. Thus we have a way to check that the loss of heterozygosity (through the fixation index $F_{st}$) is inversely proportional to the fitness prevalence for an $i$ site in the geography:

$$
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
$$
If we consider all sites in the available geography, we call this harmonic mean of fitness *Suitability prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e. environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred. Conversely, it is expected that in populations with a high fixation index value, fitness will not have prevailed over the same time interval even though conditions currently exist and populations are present.

@@

The parameterization was performed with the default values of the program with the exception of the "extrapolate" and "clamping" options to avoid artificial extrapolations in the extreme values of the climatic variables used in the models. To calibrate the models we used 70% of the records and the remaining 30% to validate them.

@@

To test our hypothesis of genetic diversity as a function of environmental suitability prevalence, from a set of genetically structured populations, a matrix of fixation indices between populations is considered. For this work we considerer the data from @@bono2018genome. It consideres 10 populations along the known distribution and provides values of $F_st$ between this populations in matrix form. The matrix was transformed into a table as pairs between populations with index values between each pair. The values of the population with itself, i.e. $0$, were included. Subsequently, georeferences were obtained for each population considered in the fixation index matrix. From this geo-referencing the SPA value for each population was extracted. Finally, the SPA value was aligned in the table of fixation index pairs. In total the table has 4 columns: source population, target population, fixation index and SPA of the target population. This table was separated by source population groups to generate a set of linear models between SPA values and fixation indexes $F_{st}$. An analysis was performed for each source population with the hypothesis that the correlation between SPA and fixation index exists and is non-zero if a dispersal process has occurred from the historical endemic area of the taxon.

>>>>

## Abstract

@@

As a case study we consider squirrels (*Sciurus aberti*), a species currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many].

Our results reveal that suitability prevalence corresponds to the fixation index of *S. aberti* populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value. Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.

@@

For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$. We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$. Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions. If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$. i. e. the set of environmental conditions for which values of fitness are capable of sustaining populations. We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$. This is a way of expressing the fundamental niche as a function of environmental conditions. So we can correspong each point of geography to environmental space,  with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$. Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.

@@

The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $S_i(\vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

@@

![Hypothetical niche in environmental space. An arbitrary point moves within the environmental space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}

@@

Moreover, the effective population can be estimated as the harmonic mean of the population in a demographic time series [@karlin1968rates] for $\tau$ generations $\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)$ for each $i$ site in $\mathbf{G}$. We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close:

$$
F_{i}^{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_{ij}} 
$$

@@

We call this harmonic mean of environmental suitability *Suitability prevalence index *($S$) given between $0$ and $1$, since it reflects the rate of population growth as a function of environment over a period of time. Thus we have a way to check that the loss of heterozygosity (through the fixation index $F_{st}$) is inversely proportional to the suitability prevalence for an $i$ site in the geography:

$$
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
$$
If we consider all sites in the available geography, we call this harmonic  *Suitability Prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e. environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred. Conversely, it is expected that in populations with a high fixation index value, fitness will not have prevailed over the same time interval even though conditions currently exist and populations are present.

@@

The parameterization was performed with the default values of the program with the exception of the "extrapolate" and "clamping" options to avoid artificial extrapolations in the extreme values of the climatic variables used in the models. To calibrate the models we used 70\% of the records and the remaining 30\% to validate them.

@@

To test our hypothesis of genetic diversity as a function of environmental suitability prevalence, from a set of genetically structured populations, a matrix of fixation indices between populations is considered. For this work we considerer the data from @@bono2018genome. It consideres 10 populations along the known distribution and provides values of $F_{st}$ between this populations in matrix form. The matrix was transformed into a table as pairs between populations with index values between each pair. The values of the population with itself, i.e. $0$, were included. Subsequently, georeferences were obtained for each population considered in the fixation index matrix. From this geo-referencing the SPA value for each population was extracted. Finally, the SPA value was aligned in the table of fixation index pairs. In total the table has 4 columns: source population, target population, fixation index and SPA of the target population. This table was separated by source population groups to generate a set of linear models between SPA values and fixation indexes $F_{st}$. An analysis was performed for each source population with the hypothesis that the correlation between SPA and fixation index exists and is non-zero if a dispersal process has occurred from the historical endemic area of the taxon.

### 2023-07-11 manuscript_english.tex
change abstract

The objective of our approach is not to know the function of \(R_0\) but
rather to relate this local variation and the function \(\mathbf{N}_F\)
with a function \$ S\_i(~vec\{e\}\_i)\$ of environmental suitability
that can be estimated from known methods of applied mathematics and
machine learning (for example the case of Maxent, Phillips, Anderson,
and Schapire (2006)).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  model the potential distribution of a taxon with present-day
  environmental conditions and obtain an environmental suitability map.
\item
  Project the potential distribution model to various environmental
  scenarios in the past with constant time steps and obtain the
  environmental suitability for each time scenario.
\item
  Characterize past analogous climates and retain projections only for
  sites with climates analogous to the present.
\item
  Generate a harmonic mean suitability map for all time slices.
\end{enumerate}

>>>>

The objective of our approach is not to know the function of \(R_0\) but
rather to relate this local variation and the function \(\mathbf{N}_F\)
with a function \(S_i(\vec{e}_i)\) of environmental suitability that can
be estimated from known methods of applied mathematics and machine
learning (for example the case of Maxent, Phillips, Anderson, and
Schapire (2006)).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Model the potential distribution of a taxon with present-day
  environmental conditions and obtain an environmental suitability map.
\item
  Project the potential distribution model to various environmental
  scenarios in the past with constant time steps and obtain the
  environmental suitability for each time scenario.
\item
  Characterize past environments and retain projections only for sites
  with analogous environmental conditions to the present.
\item
  Calculate the harmonic mean of the suitability for all the sites all
  time slices.
\end{enumerate}

### 2023-07-11 manuscript_english_word.Rmd
change abstract

---
title: "Suitability prevalence area index in late quaternary explains genetic"
author: "Norma Hernandez, Angel Robles, Nathan Upham"
date: "2023-07-08"
output: word_document
bibliography: R21_citation.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

Our results reveal that fitness prevalence corresponds to the fixation index of *S. aberti* populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value. Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.

@@

We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables. Then $R_i(\vec{e}_i)$ is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of fitness that depends on environmental conditions.

That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.

This is a way of expressing the fundamental niche as a function of environmental conditions. So we can map each point of geography to ambient space. Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates. However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.

Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$. Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space. Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.

![Hypothetical niche in environmental space. An arbitrary point moves within the ambient space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}

@@

This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function. That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$. In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and that for each trajectory the fitness function changes and therefore the suitability function changes but not accordingly. in a homogeneous way, if not more so that each site presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

Es decir:

$$
N_i \propto  S_i(\vec{e}_i)
$$
Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$

>>>>

---
title: "Suitability prevalence area index in late quaternary explains genetic diversity"
author: "Norma Hernandez, Angel Robles, Nathan Upham"
date: "2023-07-08"
output: word_document
bibliography: R21_citation.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

Our results reveal that suitability prevalence corresponds to the fixation index of *S. aberti* populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value. Finally, this study allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.

@@

For a given region $\mathbf{G}$ in te geography, at any point in time, we have a set of existing environmental conditions $\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}$. We define $\vec{e}_i =(e_1, e_2, ..., e_v )_i$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables $\mathbf{E}(\mathbf{G}, t)$. Then $R_0(\vec{e}_i)$ is a function that relates fitness to environment, that is, for a species with non-overlapping generations, the population's net reproductive rate $R_0$ is a measure of fitness that depends on environmental conditions. If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}$. i. e. the set of environmental conditions for which values of fitness are capable of sustaining populations. We call $\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)$. This is a way of expressing the fundamental niche as a function of environmental conditions. So we can correspong each point of geography to environmental space,  with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates depending on time because we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time: $R_0 = R_0(\vec{e}(t)) \}$. Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$.

@@

This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function. That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_0(\vec{e}_i) > 1$ and $1$ when $R_0$ is evaluated at the expected value of $\vec {e}(t)$ for all values of the set $\mathbf{N}_F$ where  $R_0(\vec{e}_i) > 1$. In a practical way, instead of $\mathbf{N}_F$, we approximated
$\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F$ 
(the existing niche) from known presence points in the geography $\mathbf{G}$ obtained from emprical observations stored in database, and map to $\mathbf{E}$ in order to numerically fitting a convex volume of the environmental space that enclose the known information.

The objective of our approach is not to know the function of $R_0$ but rather to relate this local variation and the function $\mathbf{N}_F$ with a function $S_i(\vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum).

Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$
$$
N_i \propto  S_i(\vec{e}_i)
$$

![Hypothetical niche in environmental space. An arbitrary point moves within the environmental space entering and exiting the volume (niche) where the intrinsic growth rate is positive.](all_figures/figure_1.png){width=50%}

Once we have an approximation of this volume, we consider that each point in the geography is static and has a trajectory in the environmental space and for each trajectory the fitness function changes and therefore the suitability function changes but not in a homogeneous way, so each site presents different displacement rates in envrionmental spaces and as a consequence, different demographic rates. So we can establish that for each site $i$ in geographic space $\mathbf{G}$, there is a population size $N_i$ that changes with respect to a fitness function dependent on time-varying environmental conditions:

$$
N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
$$
